---
title: "Effect of stay-at-home order on Covid-19 infectious population growth rate"
subtitle: "Analysis of 5 European countries: France, Germany, UK, Italy, and Poland during October and November, 2020"
author: "Team 2: Zeng Fung Liew (913802324), Yixing Lu (915501254), Liela Meng (917843295), Apurv Srivastav (918936075)"
date: "March 5th, 2021"
output:
  html_document:
    df_print: paged
    number_sections: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# Abstract 

Coronavirus disease 2019 is a contagious disease that has led to one of the biggest pandemics across the world. There has been 115 million cases of Covid-19 with 2.5 million deaths so far. One of the responses by countries has been to issue a stay-at-home mandate which has levels that range from no restrictions to a mandatory curfew. In this study, the stay-at-home mandate with apporiate lag is used to estimate the growth rate of infectious population. The exploratory data analysis showed that the Daily Percent change in infectious population. From the result, it is seen that each of the five countries had an increasing percent change at the beginning of the time interval (October 1st), but over time the daily percent change turned negative. This distribution was the same for all five countries and possibly indicates a negative association between stay-at-home mandate and infectious population. In the inferential analysis, it shown using multiple tests that Fixed effects model was better than random effects and ordinary least squares. We then use propensity scores to test whether the significant variables in the fixed effect model can cause the growth rate of infectious population to change.

# Introduction

## Coronavirus Background
Coronavirus disease 2019 (COVID-19) is a contagious disease caused by severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2). The first case was identified in Asia in December 2019 and has rapidly spread worldwide, leading to an ongoing pandemic. The spread of COVID-19 was faster than previous pandemics like that of the influenza pandemic of 1918 because of interconnectedness of a global economy and the ease of communication that that has increased global travels exponentially. As of March 2021, there have been over 115 million cases of Covid-19 with 2.5 million deaths. [6]  Symptoms of COVID-19 are variable, but often include fever, cough, fatigue, breathing difficulties, and loss of smell and taste. SARS-CoV-2, the virus that causes COVID-19, is primarily spread when an infected person is in close contact with other people. Small droplets and aerosols that contain the virus can spread from an infected person's nose and mouth when they breathe or talk. Then the surrounding people are infected when those droplets and aerosols enter their nose and mouth. Research has shown that infection mainly happens when people are near each other for long enough. People who are infected can transmit the virus sometimes up to two days before they themselves show symptoms. It has been seen that asymptomatic victims of the Covid-19 virus can still spread the virus if in close contact with another person. The ways to prevent or reduce infection include the practice of social distancing, ventilating indoor facilities, covering one’s mouth and nose when sneezing, washing hands, and avoid touching one’s face. Another method that has been employed to reduce infection is to use a mask to cover one’s face and nose as studies have shown that this minimizes the transmission risk [7]. Studies have also shown that there is a delay between the moment a person first is infected and the appearance of the preliminary symptoms. Symptoms may begin one to fourteen days after exposure to the virus. The median delay for COVID-19 is four to five days. Most symptomatic people experience symptoms within two to seven days after exposure, and almost all will experience at least one symptom within 12 days [8]. Health professionals recommend quarantining for 10-14 days after potential exposure to the virus. <br>
Social distancing across the world has been an effective measure to prevention. The social distancing measures recommended by health professionals include online learning at both the primary and college levels. Employees e-commute to work. Bars and restaurants move from dine-in to delivery. Malls were closed and shopping for food related and non-food related items went online.  Most importantly, people were recommended to wear masks in public. Governments across the world have issued mandates ranging from strict adherence to social distancing to laxer or no mandates on social distancing. The strictest of these measures by government of some countries was to issue a Stay-at-Home mandate.  The stay-at-home mandate was based on surge in COVID cases and ranged from no restrictions to mandatory indoor curfew [9]. <br>


```{r setup, include=FALSE,include=F}
rm(list=ls())
library(tidyverse)
library(dplyr)
library(zoo)
library(gridExtra)
library(scales)
library(lubridate)
library(ggplot2)
library(plotly)
options(digits=2)
```
```{r covid_window,include=F}
covid <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
covid[which(covid$Country == "The United Kingdom"), "Country"] = "United_Kingdom"
# write_csv(covid,file="WHO-Covid-19-backup.csv")

# filter out countries in "Other" WHO regions
covid <- covid %>% 
  filter(WHO_region != "Other") %>% 
  mutate(WHO_region = fct_recode(WHO_region,
                                 "Eastern Mediterranean"="EMRO",
                                 "Europe" = "EURO",
                                 "Africa" = "AFRO",
                                 "Western Pacific" = "WPRO",
                                 "Americas"="AMRO",
                                 "South-East Asia" = "SEARO"))

# stay-at-home dataset
home =  read_csv("stay-at-home-covid.csv")            # change filename/directory if necessary
home$stay_home_requirements = as.factor(home$stay_home_requirements)
home[which(home$Entity == "United Kingdom"), "Entity"] = "United_Kingdom"

# inner join data
covid = inner_join(covid, home, 
                     by = c("Date_reported" = "Date",
                            "Country" = "Entity"),
                     keep = FALSE)
covid = covid[, c("Date_reported", 
                  "Country_code", 
                  "Country", 
                  "WHO_region", 
                  "New_cases", 
                  "Cumulative_cases", 
                  "New_deaths", 
                  "Cumulative_deaths", 
                  "stay_home_requirements")]

# Target period of analysis: 2020-10-01 to 2020-11-30
# Target country: Germany, UK, France, Italy, Poland
covid_window <- covid %>% 
  filter(Country %in% c("Germany", "United_Kingdom", "France", "Italy", "Poland") & 
           Date_reported>"2020-08-15" & 
           Date_reported<"2020-12-07")

# virus incubation period d = 7
# I_(t-d-1) = y_(t-1) - y_(t-d-1), where y is cumulative cases
# delta_I_t = [y_t - y_(t-d)] - [y_(t-1)-y_(t-d-1)] = new_cases_t - new_cases_(t-d)
covid_model.d7 <- covid_window %>% 
  group_by(Country) %>% 
  mutate(New_infectious = New_cases - lag(New_cases,7),
         lag_infectious = Cumulative_cases - lag(Cumulative_cases,7))

```
```{r covid_model.d7,include=F}
# calculating 3, 5, 7-day moving average of new_infectious and cum_infectious
covid_model.d7 <- covid_model.d7 %>% 
  group_by(Country) %>% 
  mutate(New_infec_07ma = zoo:: rollmean(New_infectious,k=7,fill=NA),
         lag_infec_07ma = zoo:: rollmean(lag_infectious,k=7,fill=NA),
         New_infec_05ma = zoo:: rollmean(New_infectious,k=5,fill=NA),
         lag_infec_05ma = zoo:: rollmean(lag_infectious,k=5,fill=NA),
         New_infec_03ma = zoo:: rollmean(New_infectious,k=3,fill=NA),
         lag_infec_03ma = zoo:: rollmean(lag_infectious,k=3,fill=NA)) %>%
  dplyr::ungroup()

covid_model.d7 <- covid_model.d7 %>% 
  filter(covid_model.d7$Date_reported>"2020-09-30" & 
           covid_model.d7$Date_reported<"2020-12-01")
range(covid_model.d7$Date_reported) # 2020-10-01 to 2020-11-30

# Calculate response variable: delta_infectious
covid_model.d7 <- covid_model.d7 %>% 
  mutate(delta_infectious = 100*New_infectious/lag_infectious, 
         delta_infec_07ma = 100*New_infec_07ma/lag_infec_07ma, 
         delta_infec_05ma = 100*New_infec_05ma/lag_infec_05ma, 
         delta_infec_03ma = 100*New_infec_03ma/lag_infec_03ma)
```
```{r liela,include=F}
liela <- covid_model.d7 %>% select(Date_reported,Country,New_cases,Cumulative_cases,New_deaths,Cumulative_deaths,stay_home_requirements,delta_infectious,delta_infec_07ma,stay_home_requirements) %>% mutate(rate=New_deaths/Cumulative_cases)
# New_infectious: Infectious population is defined as people who are infected and have the potential to infect others.
# delta_infectious: percent change in infectious population on day t
```

```{r world daily - setup,include=F}
# abc <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
abc <- covid %>%  mutate(rate=New_deaths/Cumulative_cases, month=lubridate::month(Date_reported)) %>% select(-Country_code, WHO_region) # %>% dplyr::filter(!is.na(rate)) 
world <- abc %>% group_by(Date_reported) %>% summarise(newdeath=mean(New_deaths),newcase=mean(New_cases)) %>% mutate(month=lubridate::month(Date_reported)) %>% mutate(month=as.factor(month)) 
rate <- abc %>% dplyr::filter(!is.na(rate)) %>% group_by(Date_reported) %>% summarise(case_mortality=mean(rate))%>% mutate(month=lubridate::month(Date_reported)) %>% mutate(month=as.factor(month)) 
# plotting

death <- world %>% plot_ly(x=~Date_reported,y=~newdeath,type = 'bar',color=~month)%>% 
  layout(
         xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         yaxis = list(title = "daily new death",
                      zeroline = FALSE))
case <- world %>% plot_ly(x=~Date_reported,y=~newcase,type = 'bar',color=~month)%>% 
  layout(
         xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         yaxis = list(title = "daily new case",
                      zeroline = FALSE))
case_rate <- rate %>% plot_ly(x=~Date_reported,y=~case_mortality,type = 'bar',color=~month)%>% 
  layout(
         xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         yaxis = list(title = "daily case mortality",
                      zeroline = FALSE))
```
```{r world daily,echo=FALSE,fig.height=5 , fig.width=10}
subplot( case, death, case_rate, nrows = 3,shareX = T, shareY = F, titleX = T, titleY = T) %>% layout(title = "Figure 1: World average daily new case/new death/case mortality", showlegend=FALSE,showlegend2=FALSE)
```

From figure 1, the world average daily new case/new death/case mortality, we can see that though the daily case mortality remains at a steady low level after June, the daily new death has a jump from Oct to Nov, and the daily new case has a sharp increase in October. Besides, since many countries announced the stay-at-home policy between Oct and Nov, we would narrow the time of interest into this period - October and November.

```{r continent setup,include=F}
continent <- covid %>% 
  filter(WHO_region != "Other") %>% 
  mutate(WHO_region = fct_recode(WHO_region,
                                 "Eastern Mediterranean"="EMRO",
                                 "Europe" = "EURO",
                                 "Africa" = "AFRO",
                                 "Western Pacific" = "WPRO",
                                 "Americas"="AMRO",
                                 "South-East Asia" = "SEARO"))%>% mutate(WHO_region=as.factor(WHO_region))%>%  group_by(WHO_region,Date_reported) %>% summarise(newdeath=mean(New_deaths),newcase=mean(New_cases)) 

con_death <- continent  %>% plot_ly(x=~Date_reported,y=~newdeath,color = ~WHO_region) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "daily new death",
                      zeroline = FALSE)) 

con_case <- continent  %>% plot_ly(x=~Date_reported,y=~newcase,color = ~WHO_region) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "daily new case",
                      zeroline = FALSE)) 
```
```{r continent,echo=FALSE,fig.height=5 , fig.width=10}
subplot( con_case,con_death, nrows = 2,shareX = T, shareY = F, titleX = T, titleY = T) %>% layout(title = "Figure 2: Average daily new case/new death of each WHO Region",showlegend=FALSE,showlegend2=FALSE)

```

Figure 2 shows the average daily new case/new death of each WHO Region. For different regions, the daily new cases or new death are relatively different. To gain a more robust result, we would focus our analysis on one region rather than the whole world. Thus, region Europe was selected, and countries that fit our inclusion criteria were included in this analysis. To sum up, we narrow our analysis to five countries - UK, Poland, France, Italy, Germany between October and November. <br>

The five European countries were identified as being matched on comorbidities as well as similarity on transmission based on lower variability in enforcing and implementing mandates on social distancing. Another reason these five countries were chosen because of similarity on demographics like race and ethnicity [10]. Both of which have been shown to have an effect on Covid-19. Furthermore, data from October 1st to November 30th was selected based on the distribution of similarity in COVID-19 cases, as well as the non-availability of COVID-19 vaccines during this period for these countries. Further, there were fewer mutations of the virus till this timepoint [11].<br>
There are two primary motivations of this project. <br>

## Obejctive
There were two major objectives of the current project:

(1) **Find the lagged effect of stay-at-home order.** It is common that a policy takes time to reveal its effect after implementation. The lagged time of stay-at-home order on Covid-19 infectious population growth rate was unclear. Thus, we constructed a model to find the number of days after the policy started when its effect first became significant.

(2) **Investigate the effect of strictness of stay-at-home order on growth rate of infectious population.** Using the lagged time found in the first part, this model would investigate how the strictness of policy and change in such policy affect the response variable. 

 
# Background 
## Target population
During October and November 2020, daily reported cases in Americas and Europe skyrocketed. In response to such rapid increase in infected cases, many countries strengthened their stay-at-home order as one of the acts to control the spread of Covid-19. The question of interest is the effect of such policy response on containing the spread of disease. Admittedly, there are many factors other than the stay-at-home order that could affect the spreading rate, especially there were big regional differences among countries across the world, such as weather, economic status, demographics, medical conditions, and other Covid-19 related policies. Also, not all the regions had such peaks in Covid-19 cases as observed in Americas and Europe. In order to establish a causal inference of stay-at-home order on Covid-19 spread, we would like to target countries with similar geographic locations, and experienced similar total number and trends of Covid-19 cases during October to November 2020. Given such consideration, we decided to look at the top five European countries with population density over 100/km$^2$ and land area over 100,000 mi$^2$. Since we were concerned about the spread of Covid-19, it was reasonable to focus on countries with comparable population density as well as total number of cases. _Using such criteria, the five countries we focused our analysis on are: Germany, UK, France, Italy, and Poland._ All five countries made some changes to their stay-at-home policy during the period between October 1 and November 31,2020.

## Data
Contry level daily and cumulative Covid-19 cases and deaths data was obtained from the World Health Organization (WHO) website (https://covid19.who.int/WHO-COVID-19-global-data.csv). The country level stay-at-home order information was obtained from "Our World in Data - statistics and research on conronavirus pandemmic" (https://ourworldindata.org/covid-stay-home-restrictions). The source of the data was published by Oxford COVID-19 government response tracker and was maintained up-to-date. 

### Predictors
The predictor "stay-at-home order" is a categorical variable with four levels:

* 0 - no measures
* 1 - recommend not leaving house
* 2 -  require not leaving house with exceptions for daily exercise, grocery shopping, and ‘essential’ trips
* 3 - Require not leaving house with minimal exceptions (e.g. allowed to leave only once every few days, or only one person can leave at a time, etc.)

None of the country in our target population had the most strict level "3" measure.

### Response variable
The spread of a disease is an exponential process. The classic SIR model in epidemiology describes the process as follows: in a population (N), each of the infected ones (I) had the potential to infect one or more susceptible people (S) before they recover or die (R). At any time $t$, we have $N = I(t) + S(t) + R(t)$, assuming people who have recovered will be immuned. Based on this model, for a country c at time t, the growth rate of infectious population is defined as: $$dI_{ct}/dt = I_{ct} - I_{c,t-1} = \beta I_{c,t-1} \frac{S_ct}{N_ct} - \gamma I_{c,t-1}$$
where $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery. Since the total number of people who have been infected was small compared to the whole population, we assumed that $\frac{S_ct}{N_ct} =1$. 

The WHO dataset had information on daily new cases and cumulative cases, but not the number of recoveries, therefore we could not know the exact number of infectious population (I). In order to estimate it, we assumed that it took 7 days for a person to recover from Covid-19, starting from the day it was reported. That is, people who were reported to be positive on day 0 would be removed from the infectious population on day 7. Such an estimate could be not that accurate due to lack of information. Adjutment of this parameter could be done in future study. With this assumption, infectious population could be estimated as:
\[
\begin{aligned}
I_{ct} &= y_{ct} - y_{c,t-7} \\
I_{c,t-1} &= y_{c,t-1} - y_{c,t-8}
\end{aligned}
\]
where $y_{ct}$ is the number of cumulative case in country c on date t.

With this estimate, the growth rate of infectious population (percentage point)

$$\Delta y_{ct} = \frac{I_{ct} - I_{c,t-1}}{I_{c,t-1}} = \frac{(y_{ct} - y_{c(t-7)}) - (y_{c(t-1)} - y_{c(t-8)})}{y_{c(t-1)} - y_{c(t-8)}}$$
was used as the response variable to reflect the spread of Covid-19.

# Descriptive analysis 
Our question of interest is related to stay at home policy. Hence, we would like to see what happens after the policy was announced. 

```{r new death/cases-setup, include=FALSE}
France <- liela %>% filter(Country=="France")  
United_Kingdom <- liela %>% filter(Country=="United_Kingdom")  
Germany <- liela %>% filter(Country=="Germany")  
Italy <- liela %>% filter(Country=="Italy")  
Poland <- liela %>% filter(Country=="Poland")  
# new death
F_bar <- France %>% plot_ly(x=~Date_reported,y=~New_deaths,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "France",
                      zeroline = FALSE)) 
UK_bar <- United_Kingdom %>% plot_ly(x=~Date_reported,y=~New_deaths,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "UK",
                      zeroline = FALSE)) 
G_bar <- Germany %>% plot_ly(x=~Date_reported,y=~New_deaths,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Germany",
                      zeroline = FALSE)) 
I_bar <- Italy %>% plot_ly(x=~Date_reported,y=~New_deaths,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Italy",
                      zeroline = FALSE)) 
P_bar <- Poland %>% plot_ly(x=~Date_reported,y=~New_deaths,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Poland",
                      zeroline = FALSE)) 
# new case
F_barc <- France %>% plot_ly(x=~Date_reported,y=~New_cases,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "France",
                      zeroline = FALSE)) 
UK_barc <- United_Kingdom %>% plot_ly(x=~Date_reported,y=~New_cases,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "UK",
                      zeroline = FALSE)) 
G_barc <- Germany %>% plot_ly(x=~Date_reported,y=~New_cases,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Germany",
                      zeroline = FALSE)) 
I_barc <- Italy %>% plot_ly(x=~Date_reported,y=~New_cases,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Italy",
                      zeroline = FALSE)) 
P_barc <- Poland %>% plot_ly(x=~Date_reported,y=~New_cases,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Poland",
                      zeroline = FALSE)) 
```
```{r new death/cases,echo=FALSE,fig.height=5 , fig.width=10}
subplot( F_bar, UK_bar, G_bar, I_bar, P_bar, nrows = 5,shareX = T, shareY = F, titleX = T,titleY=T) %>% layout(title = "Figure 3: Daily new death of each country",showlegend=FALSE,showlegend2=FALSE)
subplot( F_barc, UK_barc, G_barc, I_barc, P_barc, nrows = 5,shareX = T, shareY = F, titleX = T,titleY=T) %>% layout(title = "Figure 4: Daily new cases of each country",showlegend=FALSE,showlegend2=FALSE)

```
```{r pie chart-setup,include=F}
France_pie <- liela %>% filter(Country=="France") %>% mutate(stay_home_requirements=as.integer(stay_home_requirements)) %>% group_by(stay_home_requirements) %>% summarise(n=n()) %>%  mutate(policy = recode(stay_home_requirements,
  `0` = "No measures",
  `1` = "recommend not leaving house",
  `2` = "require not leaving house with exceptions for daily exercise, grocery shopping, and ‘essential’ trips",
  `3` = "Require not leaving house with minimal exceptions (e.g. allowed to leave only once every few days, or only one person can leave at a time"
))
United_Kingdom_pie <- liela %>% filter(Country=="United_Kingdom")  %>% mutate(stay_home_requirements=as.integer(stay_home_requirements)) %>% group_by(stay_home_requirements) %>% summarise(n=n()) %>%  mutate(policy = recode(stay_home_requirements,
  `0` = "No measures",
  `1` = "recommend not leaving house",
  `2` = "require not leaving house with exceptions for daily exercise, grocery shopping, and ‘essential’ trips",
  `3` = "Require not leaving house with minimal exceptions (e.g. allowed to leave only once every few days, or only one person can leave at a time"
))
Germany_pie <- liela %>% filter(Country=="Germany")  %>% mutate(stay_home_requirements=as.integer(stay_home_requirements)) %>% group_by(stay_home_requirements) %>% summarise(n=n()) %>%  mutate(policy = recode(stay_home_requirements,
  `0` = "No measures",
  `1` = "recommend not leaving house",
  `2` = "require not leaving house with exceptions for daily exercise, grocery shopping, and ‘essential’ trips",
  `3` = "Require not leaving house with minimal exceptions (e.g. allowed to leave only once every few days, or only one person can leave at a time"
))
Italy_pie <- liela %>% filter(Country=="Italy")  %>% mutate(stay_home_requirements=as.integer(stay_home_requirements)) %>% group_by(stay_home_requirements) %>% summarise(n=n()) %>%  mutate(policy = recode(stay_home_requirements,
  `0` = "No measures",
  `1` = "recommend not leaving house",
  `2` = "require not leaving house with exceptions for daily exercise, grocery shopping, and ‘essential’ trips",
  `3` = "Require not leaving house with minimal exceptions (e.g. allowed to leave only once every few days, or only one person can leave at a time"
))
Poland_pie <- liela %>% filter(Country=="Poland")%>% mutate(stay_home_requirements=as.integer(stay_home_requirements)) %>% group_by(stay_home_requirements) %>% summarise(n=n()) %>%  mutate(policy = recode(stay_home_requirements,
  `0` = "No measures",
  `1` = "recommend not leaving house",
  `2` = "require not leaving house with exceptions for daily exercise, grocery shopping, and ‘essential’ trips",
  `3` = "Require not leaving house with minimal exceptions (e.g. allowed to leave only once every few days, or only one person can leave at a time"
))

fig <- plot_ly()
fig <- fig %>% add_pie(data = France_pie, labels = ~policy, values = ~stay_home_requirements,
          name = "France", domain = list(x = c(0, 0.3), y = c(0.4, 0.7)))
fig <- fig %>% add_pie(data = Italy_pie, labels = ~policy, values = ~stay_home_requirements,
          name = "Italy", domain = list(x = c(0.4, 1), y = c(0.4, 0.7)))
fig <- fig %>% add_pie(data =Poland_pie, labels = ~policy, values = ~stay_home_requirements,
          name = "Poland", domain = list(x = c(0, 0.3), y = c(0, 0.3)))
fig <- fig %>% add_pie(data =Germany_pie, labels = ~policy, values = ~stay_home_requirements,
          name = "Germany", domain = list(x = c(0.4, 0.7), y = c(0, 0.3)))
fig <- fig %>% add_pie(data =United_Kingdom_pie, labels = ~policy, values = ~stay_home_requirements,
          name = "UK", domain = list(x = c(0.7,1), y = c(0, 0.3)))
fig <- fig %>% layout(title = "Figure 5:Pie Charts of stay-at-home requirements", showlegend = T,shareY = F,titleY=T,
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = T),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = T),
         annotations = list(x = c(.08, .62, .08, .5, 0.8),
                            y = c(.8, .8, .3, .3,0.3),
                            text = c("France","Italy","Poland","Germany","UK"),
                            xref = "papper",
                            yref = "papper",
                            showarrow = F
                          ))
```
```{r pie chart,echo=FALSE,fig.height=5 , fig.width=10}
fig
```


From figures 3 and 4, the daily new deaths/cases of each country, we can barely see the impact of stay-at-home policy while looking at daily new cases or daily new death for countries and time we selected. We can observe that daily new death/case increases while after the policy was published. Thus, to investigate whether the policy works, instead of daily new cases or deaths, we can look at the percent change in infectious populations (delta infectious). Figuer 5 shows stay-at-home plicy change for all five countries. Figure 6 shows the daily delta infectious for each country, and this delta infectious decrease in general and were negative after Nov 15, which indicates quarantine had an impact. 

```{r delta_infectious-setup, include=F}
Fi_bar <- France %>% plot_ly(x=~Date_reported,y=~delta_infectious,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "France",
                      zeroline = FALSE)) 
UKi_bar <- United_Kingdom %>% plot_ly(x=~Date_reported,y=~delta_infectious,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "UK",
                      zeroline = FALSE)) 
Gi_bar <- Germany %>% plot_ly(x=~Date_reported,y=~delta_infectious,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Germany",
                      zeroline = FALSE)) 
Ii_bar <- Italy %>% plot_ly(x=~Date_reported,y=~delta_infectious,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Italy",
                      zeroline = FALSE)) 
Pi_bar <- Poland %>% plot_ly(x=~Date_reported,y=~delta_infectious,color = ~stay_home_requirements) %>%
  add_bars() %>%
  layout(xaxis = list(title = "Date reported",
                      zeroline = FALSE),
         barmode = "stack",
         yaxis = list(title = "Poland",
                      zeroline = FALSE))
```
```{r delta_infectious,echo=FALSE,fig.height=5 , fig.width=10}
subplot( Fi_bar, UKi_bar, Gi_bar, Ii_bar, Pi_bar, nrows = 5,shareX = T, shareY = F, titleX = T,titleY=T) %>% layout(title = "Figure 6: Daily percent change in infectious populatio for each country",showlegend=FALSE,showlegend2=FALSE)
```

# Inferential analysis 

## Lagged effect of stay-at-home order 

```{r set up, include = FALSE}
library(tidyverse)
library(dplyr)
library(zoo)
library(ggplot2)
library(plotly)
library(plm)
library(knitr)
library(broom)
# import data
# covid data

covid_model.d7 = read.csv("who_new_response.csv", header = TRUE)
# stay-at-home data

```
```{r pre-processing}
# data pre-processing
# filter data
home <- home %>% filter(Entity %in% c("Germany", "United_Kingdom", "France", "Italy", "Poland") & Date>"2020-09-30" & Date<"2020-12-01")
# find the date that each country tightened their stay-at-home order
ind = home %>% group_by(Entity) %>% mutate(change= stay_home_requirements-lag(stay_home_requirements,1))
start_date = ind[which(ind$change!=0),]
# response variable_by_days_since_policy_started
covid_new_d7 = covid_model.d7
# the date that stay-at-home order was tightened
policy_start = rep(0,dim(covid_new_d7)[1]) 
policy_start[which(covid_new_d7$Country=="France")] = "2020-10-17" 
policy_start[which(covid_new_d7$Country=="Germany")] = "2020-10-15" 
policy_start[which(covid_new_d7$Country=="Italy")] = "2020-10-23"
policy_start[which(covid_new_d7$Country=="Poland")] = "2020-10-24"
policy_start[which(covid_new_d7$Country=="UK")] = "2020-10-23"
covid_new_d7$policy_start = as.Date(policy_start)
covid_new_d7$Date_reported = as.Date(covid_new_d7$Date_reported)
# number of days prior to or after policy issued
days_since_start = rep(0, dim(covid_new_d7)[1])
days_since_start = covid_new_d7$Date_reported - covid_new_d7$policy_start
covid_new_d7$days_since_start = days_since_start
# new rsponse dataset by days since policy started
covid_by_days = covid_new_d7[which(covid_new_d7$days_since_start>-15 & covid_new_d7$days_since_start<29),]
```

```{r by-days plot, fig.height=4, fig.width=10}
# plots of growth of infectious population vs. days since order started
covid_by_days$days_since_start <- as.factor(covid_by_days$days_since_start)
infection_country_bydays = covid_by_days %>% ggplot(aes(x=days_since_start,y=delta_infec_07ma, group=Country,color=Country))+geom_line()+labs(x = "Days since start of stay-at-home order", y = "Growth rate of infectous population")+ggtitle("Growth rate of infectious population before and after the start of stay-at-home order") +theme(plot.title = element_text(size = 11,face = "bold"),legend.text = element_text(size=10))
ggplotly(infection_country_bydays,tooltip = c("x","colour","y"))
```

### Model justification
As observed in the world trend of daily cases of Covid-19, there was a peak between October and November in many countries, including the five European countries that we focused on in this project. In response to such skyrocketing numbers of cases, all five contries issued stay-at-home orders around that period of time in hope to contain the spread of Covid-19. It was also noticed that there was a delay between the order starting date and the date when daily new cases began to decrease, which was expected since most of policies would have a lagged effect. _In order to understand the temporal dynamics of the stay-at-home order, we measured the policy effects on growth rate of infectious population as a function of time since the order started._ Number of days since the order started was treated as a categorical variable with each day as a level. We looked at the period of 14 days before and 28 days after the order came into effect. The estiamted coefficients could be considered as the change in infectious population growth rate due to the stay-at-home order without varying implementation time.

A fixed effect regression model was constructed. One advantage of this model is that it controls for unobserved country-specific factors that were not included as predictors.This model is equivalent to a difference-in-difference model with variation in treatment time.

### Model setup and parameter notation
An one-way fixed effect regression model was defined as: $$\Delta y_{ct} = \beta_c + \sum_{\tau \neq -1}\beta_\tau X_{ct\tau}+\epsilon_{ct}$$

**Parameter notation** <br>

* $Y_{ct}$ is the response variable: % growth rate in infectous population of country c at time t.

* $\beta_c$ is the counry-specific, time-invariant fixed factor for country c.

* $\tau$ is the number of days prior to or after the stay-at-home order was issued. 

* $\beta_\tau$ is the effect size of the policy on the growth rate of infectious population $\tau$ days before/after the policy started. 

* $X_{ct\tau}$ are dummpy variables indicating whether day t is the $\tau$ day before or after policy started in country c. 

**Assumptions** <br>

* Conditional relationship of $Y_{ct}$ given $X_{ct\tau}$ is linear in the explanatory variables.

* $\epsilon_{ct}$ are independent random variables with zero mean and constant variance: $$E(\epsilon_{it})=0,\  Var(\epsilon_{it})=\sigma^2$$

**Hypothesis** <br>

* $H_0:$ all $\beta_\tau = 0$ vs. $H_a:$ not all $\beta_\tau = 0$

### Model fitting results
```{r model fitting, include = FALSE}
covid_by_days$days_since_start <- relevel(covid_by_days$days_since_start,ref = "-1")
# OLS model
ols <- lm(delta_infec_07ma~days_since_start,data = covid_by_days)
# country-fixed effect model
country_plm <- plm(delta_infec_07ma~days_since_start,data = covid_by_days,index = c("Country","days_since_start"), model = "within")
# confidence interval for each coefficient
coef = country_plm$coefficients
se = sqrt(diag(country_plm$vcov))
alpha = 0.05
cval = qt(alpha/2, country_plm$df.residual,lower.tail = FALSE)
CI_lower = coef - cval*se
CI_upper = coef + cval*se
inference <- data.frame(Days_since_start = c(-14:-2,0:28), Estimate = coef, CI_lower = CI_lower, CI_upper = CI_upper)
summary = inference[14:42,]
summary$CI_lower = round(summary$CI_lower,digits = 2); summary$CI_upper = round(summary$CI_upper,digits = 2); summary$Estimate = round(summary$Estimate,digits=2)
# R2 and Ra2
n = dim(covid_by_days)[1]
sst = with(covid_by_days, sum((delta_infec_07ma-mean(delta_infec_07ma))^2))
sse = sum(residuals(country_plm)^2)
R2 = 1 - sse/sst
Ra2 = 1- (1-R2)*(n-1)/(n-length(coef(country_plm))-1)
```

**Table 1** summarized the coefficient estimates and their 95% confidence intervals for 0 ~ 28 days after the policy started, which could be interpreted as the effect of different implemetation date on growth rate of infectious population. Negative values indicated that the growth rate of infectious population was slowing down. As shown in the table, the effect of stay-at-home order on flattering the curve started to become significant (p<0.05) at day 13 after it was first implemented. The figure below visualized the effect of stay-at-home order before and after it was implemented. The days prior to policy implementation was also included in the model to test for any reversed causal effect. The figure showed that the coefficients estimates before the policy started were not significantly different from zero, confirming that the growth rate in infectiou population did not affect the policy starting date.

```{r model results, fig.height=4,fig.width = 10, warning = FALSE}
# table of beta estimates and ci 
kable(summary,row.names = FALSE,align = "c", caption = "Table 1: Coefficient estimates from regression of daily growth rate of Covid-19 infectious population on number of days since stay-at-home order") 
# plot coefficient estimates and confidence bands
ggplot(inference, aes(x=Days_since_start,y=Estimate)) + 
  geom_line() + geom_point() + 
  geom_ribbon(mapping = aes(ymin = CI_lower, ymax = CI_upper),alpha = 0.4) + 
  labs(x = "Days since stay-at-home order",y = "\u0394 Growth in infectious population")+ ggtitle("Change in growth rate of infectious population with varying implementation time of stay-at-home order") + theme(plot.title = element_text(size = 12, face="bold")) +
  geom_vline(xintercept = c(0,13), linetype = "dashed",color = "red")
```

* The dashed red lines dictated D-day of the policy and 13 days after its implementation when its effect on reducing growth rate of infectious population firstly became significant.



## Effect of policy strictness on infectious population growth rate

To find out whether the change in level of stay-home restriction has an impact on the rate of growth in active infections, we fit a panel regression model as follows:
$$
\Delta y_{ct} = \alpha_c + \alpha_t + \beta_{1,1} x_{1,1,ct} + \beta_{1,2} x_{1,2,ct} + \beta_{2,-1} x_{2,-1,ct} + \beta_{2,1} x_{2,1,ct} + \beta_{2,2} x_{2,2,ct} + \beta_{3} x_{3,ct} + u_{ct}
$$
where 

(1) $\Delta y_{ct}$ is the change in rate of active infections for country $c$ at time $t$, which is calculated using 

$$
\Delta y_{ct} = \frac{(y_{ct} - y_{c(t-7)}) - (y_{c(t-1)} - y_{c(t-8)})}{y_{c(t-1)} - y_{c(t-8)}}
$$
where $y_{ct}$ is the number of new cases for country $c$ at time $t$.

(2) $\alpha_c$ is the country-level fixed effect, ie. $c = \{$France, Germany, UK, Poland, Italy$\}$.

(3) $\alpha_t$ is the time-level fixed effect, ie. the dates reported for this data analysis, $t = \{10/01/2020, \ldots, 11/30/2020\}$.

(4) $x_{1,i,ct}$ is the stay-home requirement level for country $c$ at 14 days prior to time $t$, ie. $i = \{0,1,2\}$. Since $x_{1,ct}$ is represented as dummy variables, the case for $x_{1,0,ct}$ is implied by the model when $x_{1,1,ct} = x_{1,2,ct} = 0$.

(5) $x_{2,j,ct}$ is the change in stay-home requirement for country $c$ at the previous week of (4). This is computed as follows:
$$
x_{2,j,ct} = \mathbb{1}_{\{x_{1,i,ct} - x_{1,i',c(t-7)}\}} \ {\rm for } \ i,i' = \{0,1,2\}, j = \{-1, 0, 1, 2\} 
$$
where $j = -1$ represents a 1 level relaxation of the stay-at-home orders, $j=0$ represents no change in the stay-at-home orders, while $j=1,2$ represents an increase in strictness of stay-at-home policy by 1 and 2 levels respectively. Since $x_{2,ct}$ is represented as dummy variables, the case for $x_{2,0,ct}$ is implied by the model when $x_{2,1,ct} = x_{2,2,ct} = x_{2,-1,ct} = 0$.

(6) $x_{3,ct}$ is the change in rate of active infections for country $c$ 14 days prior to time $t$, which is a 14-day lag of same measurement of the response variable $\Delta y_{ct}$.

(7) $u_{ct}$ is the error for country $c$ at time $t$.

This model comes with a few advantages. The fixed effect $\alpha_c$ controls the time-invariant effect of countries which can impact the change in rate of active infections for each country, eg. weather, socioeconomy status, local health care system, and population size [1]. The fixed effect $\alpha_t$ controls the factors that vary over time. As discussed by Fowler et al., this includes the changes in testing availability, national policies that vary over time, and the occurence of major events that impacts social behaviors of people in the country. However, the panel regression model also comes with some strict assumptions. These assumptions and diagnostics are tested and discussed under the Sensitivity Analysis section.

The fitted model gives us the following estimates:
```{r}
library(plm)
covid_model.d7$lag_stay_home_diff_07 = as.factor(covid_model.d7$lag_stay_home_diff_07)
covid_model.d7$lag_stay_home = as.factor(covid_model.d7$lag_stay_home)
model = plm(delta_infec_07ma ~ lag_stay_home + lag_stay_home_diff_07 +lag_delta_infec_07ma, data = covid_model.d7,
                index = c("Country", "Date_reported"), model = "within")
tt = as.data.frame(summary(model)$coef)
rownames(tt) = c("Stay-home restriction at t-14 == 1",
                 "Stay-home restriction at t-14 == 2",
                 "Change in stay-home restriction at t-14 == 0",
                 "Change in stay-home restriction at t-14 == +1",
                 "Change in stay-home restriction at t-14 == +2",
                 "Change in rate of active infections at t-14")
tt
```

The fixed effect coefficients for the countries are obtained as shown below:
```{r}
tt = as.data.frame(fixef(model))
colnames(tt) = expression(alpha[c])
tt
```

We can interpret the $\hat \alpha_c$ values above as a growth rate in active infections when there were stay-at-home policies were relaxed to level 0 in the previous 2-3 weeks. As we can see that most $\hat \alpha_c$ values are positive (except France), it means that the growth rate of active infections increases significantly in these 4 countries when there are no stay-at-home policies in place. This is especially the case in the UK with its growth rate of more than 4. 

Based on the results above, it is also obvious to see that a sticter stay-at-home policy results in a larger decrease in rate of growth in active infections. This linear association is accompanied by a strong $R^2$ and adjusted $R^2$ values of 0.71 and 0.7 respectively. Assuming there are no changes in stay-at-home policies, a level-2 stay-at-home restriction (-8.43) reduces the growth rate of active infections by more than 2 as compared to a level-1 stay-at-home restriction (-6.26).

On the other hand, a change in policy on the stay-at-home restriction 2-3 weeks prior also has a significant impact on the growth rate in active infections. Assuming a country increased its stay-at-home level within the past 2-3 weeks (eg. from level 0 to level 2), we still expect to observe a decreasing growth rate in active infections. For instance, Italy, who wsn't implementing any form of stay-at-home restriction up until mid-October, would expect to observe a decrease in growth rate of active infections by $-1.12$(ie. $- 8.43 + 7.31$) in November.

Meanwhile, it is also important to note that the change in rate of active infections are also affected by its own value from 2 weeks prior. Its coefficient estimate of 0.36 suggests that a country's change in rate of active infections will continuously increase if no stay-at-home policies were put in place.

### Siginificance of predictor variables
An important aspect when interpreting the fitted model above is to ensure that the coefficient estimates are significant. In other words, the standard errors of the coefficient estimates should not be so large such that there are no significant evidence that the predictor variables correlate with the change in rate of active infections. To test this, we will use the $F$-test to check the significance of each predictor variable.

* **14-day lagged effect of stay-at-home order ** <br>
The null and alternate hypothesis are
$$
H_0: \beta_{1,1} = \beta_{1,2} = 0 {\rm \ vs \ } H_a: \text{at least one of }\beta_{1,i} \neq 0
$$
Since the p-value obtained from the $F$-test is less than 2e-12, we reject the null hypothesis and conclude that the 14-day lagged effect of stay-at-home order is significant.
```{r, include = FALSE}
model.no_order = plm(delta_infec_07ma ~ lag_stay_home_diff_07 +lag_delta_infec_07ma, data = covid_model.d7,
                     index = c("Country", "Date_reported"), model = "within")
pFtest(model, model.no_order)       # --- lag stay home req. is significant
```

* **14-day lagged effect of change in stay-at-home order** <br>
The null and alternate hypothesis are
$$
H_0: \beta_{2,1} = \beta_{2,2} = \beta_{2,-1} = 0 {\rm \ vs \ } H_a: \text{at least one of }\beta_{2,j} \neq 0
$$
Since the p-value obtained from the $F$-test is approximately 1e-10 (< 0.001), we reject the null hypothesis and conclude that the 14-day lagged effect stay-at-home policy change is significant.
```{r, include = FALSE}
model.no_polchg = plm(delta_infec_07ma ~ lag_stay_home +lag_delta_infec_07ma, data = covid_model.d7,
                     index = c("Country", "Date_reported"), model = "within")
pFtest(model, model.no_polchg)      # --- lag policy change effect is significant
```


* **14-day lagged effect of change in rate of active infections** <br>
The null and alternate hypothesis are 
$$H_0: \beta_{3} = 0 {\rm \ vs \ } H_a: \beta_{3} \neq 0$$
Since the p-value obtained from the $F$-test is approximately 1e-12 (< 0.001), we reject the null hypothesis and conclude that the 14-day lagged effect of the change in rate of active infections is significant.
```{r, include=FALSE}
model.no_lagy = plm(delta_infec_07ma ~ lag_stay_home + lag_stay_home_diff_07, data = covid_model.d7,
                     index = c("Country", "Date_reported"), model = "within")
pFtest(model, model.no_lagy)      # --- lag y effect is significant
```

Overall, we have shown that all three predictor variables are significant in determining the change in rate of active infections.





# Sensitivity analysis 

## Lagged effect of stay-at-home order 
```{r, include = FALSE}
## model justification
# test for individual effect
plm_pooling = plm(delta_infec_07ma~days_since_start,data = covid_by_days,index = c("Country","days_since_start"), model = "pooling")
pFtest(country_plm, plm_pooling) #-------- p<2.2e-16, indicating significant individual effect
```

First, the necessity of adopting a individual fixed effect model was tested by comparing the current model with a pooling model where all countries' data were pooled together. The null hypothesis that the two models are indistinguishable was rejected (p<0.05), indicating that it was necessary to use a fixed effect model accounting for country-specific factors.

The calculated $R^2$ and adjusted $R^2$ for the model were 0.74 and 0.68 respectively. 

The diagnostics plots below tested the homocedasticity and normality assumption, as well as the presence of outliers. The residuals vs. fitted value plot showed that the residuals were mostly evenly distributed around zero but with a slightly quadratic pattern. QQ plot confirmed the normality assumption was fulfilled. And the Cook's plot showed that no observation had cook's distance larger than 1, indicating that there was no influential cases.

```{r,warning=FALSE}
## model diagnostics
###### Test for homoscedasticity ######
par(mfrow=c(2,2),oma=c(0,0,2,0))
fitted <- as.numeric(covid_by_days$delta_infec_07ma - residuals(country_plm))
# residual vs. fitted plot
plot(fitted,residuals(country_plm),bty='n',xlab='Fitted values', ylab='Residuals',main='Residuals vs. Fitted values')
lines(smooth.spline(fitted,residuals(country_plm), spar=1.3), col=2)
abline(h=0,col='red',lty='dashed')
###### Test for normality #######
qqnorm(residuals(country_plm), main = "QQ plot of residuals")
qqline(residuals(country_plm), col = "red")
###### Cook's distance ######
x.tmp <- diag(1,43,43)
x.tmp[14,14] = 0
x.tmp <-  x.tmp[,-14]
X <- rbind(x.tmp,x.tmp,x.tmp,x.tmp,x.tmp)
H <- X %*% solve(t(X) %*% X) %*% t(X)
cooks = residuals(country_plm)^2/length(coef(country_plm))* diag(H)/(1-diag(H))
plot(names(cooks),cooks,type = "h", xlab="Index",ylab = "Cook's distance", main = "Cook's distance")
points(names(cooks),cooks,cex=0.3)
title(main = "Diagnostics plots", outer=TRUE)
par(mfrow=c(1,1))
```

## Effect of policy strictness on infectious population growth rate
### Model Assumptions and Justifications
The assumptions for this model are as follows:

(1) All $x_{1,ct}$ and $x_{2,ct}$ are identically and independently distributed (i.i.d).

(2) Perfect multicollinearity within predictor variables do not exist.

(3) The error term $u_{ct}$ has conditional mean zero.

(4) Homocedasticity (constant variance) of the response variable $\Delta y_{ct}$.

(5) Influential outliers do not exist.

```{r}
par(mfrow = c(2,2))
fitted = as.vector(model$model[[1]] - model$residuals)
res = as.vector(model$residuals)
plot(fitted, res, main = "Residuals vs Fitted Values",
     xlab = "Fitted values", ylab = "Residuals")
lines(smooth.spline(fitted,res, spar=1.5), col=2)
abline(h=0,col='red',lty='dashed')          # --- homocedasticity achieved + error term has zero mean

qqnorm(res, main = "Normal Q-Q Plot of Residuals")
qqline(res)                                 # --- residuals normally distributed

x = model.matrix(model)
H = x %*% solve(t(x) %*% x) %*% t(x)
cooks = residuals(model)^2/length(coef(model))* diag(H)/(1-diag(H))
h = diag(H)
plot(h, cooks, type = "h", main = "Cook's distance vs Leverage values",
     xlab = "Leverage", ylab = "Cook's Distance")

```

Assumptions (3), (4), and (5) have been easily satisfied as shown in the plots above. In the residuals vs fitted values plot, we observe that the data is randomly distributed with no particular pattern, and that they are centered around mean zero. This satisfies the homocedasticity (constant variance) and zero mean constraint of the error term. Additionally, we also observe from the Q-Q plot that the residuals are distributed normally. The Cook's distance vs Leverage plot shows a maximum Cook's Distance value of less than 0.2, we conclude that outliers did not exist in this model. As for assumption (2), multicollinearity can be checked by calculating the variance inflation factor (VIF) of the predictor variables. Specifically, The VIF of the $k$th predictor, denoted as $VIF_k$, is defined as 
$$
VIF_k = \frac{1}{1-R_k^2}
$$
where $R_k^2$ is the coefficient of multiple determination when the predictor variable $X_k$ is regressed onto the rest of the $X$ variables. The VIF values of each predictor is shown in the table below. Since each $VIF_k$ values are close to 1, we conclude that there is no issue of multicollinearity. 
```{r}
model.pool = plm(delta_infec_07ma ~ lag_stay_home + lag_stay_home_diff_07 + lag_delta_infec_07ma, data = covid_model.d7,
               index = c("Country", "Date_reported"), model = "pooling")
tt = as.data.frame(car::vif(model.pool))
rownames(tt) = c("Stay-home restriction at t-14",
                 "Change in stay-home restriction at t-14",
                 "Change in rate of active infections at t-14")
tt                                                  # --- small VIFs, no multicollinearity
```

On the other hand, assumption (1) cannot be satisfied by this model. This is because the predictor variables such as the stay-at-home restrictions or the policy changes are not i.i.d. distributed, as each of these entity are dependent on their previous values.

### Model Diagnostics

The following are some model diagnostics that were considered for this analysis:

(1) Test for fixed effects vs random effects 
```{r, include = FALSE}
# test for ols vs fixed effects
model.ols = lm(delta_infec_07ma ~ lag_stay_home + lag_stay_home_diff_07 + lag_delta_infec_07ma, data = covid_model.d7)
pFtest(model, model.ols)                  # --- reject null, fixed-effects better

# test for ols vs random effects
plmtest(model.pool, type = c("bp"))       # --- reject null, random-effects better

# test for fixed vs random effects (requires hacky fix)
model.random = plm(delta_infec_07ma ~ lag_stay_home + lag_delta_infec_07ma, 
                   data = covid_model.d7, index = c("Country", "Date_reported"), model = "random")
model.fixed2 = plm(delta_infec_07ma ~ lag_stay_home + lag_delta_infec_07ma,
                   data = covid_model.d7, index = c("Country", "Date_reported"), model = "within")
phtest(model.fixed2, model.random)        # --- fail to reject null, random-effects better
```

(2) Test for time fixed effects
```{r, include=FALSE}
model.time = plm(delta_infec_07ma ~ lag_stay_home + lag_stay_home_diff_07 + lag_delta_infec_07ma +
                   Date_reported, data = covid_model.d7,
                 index = c("Country", "Date_reported"), model = "within")
pFtest(model.time, model)                 # --- reject null, use time-effects
plmtest(model, c("time"), type = c("bp")) # --- fail to reject null, no need for time-effects
```

(3) Test for cross-sectional dependence
```{r, include=FALSE}
pcdtest(model, test = c("lm"))        # --- there is cross sectional dependence
```

For (1), the following hypothesis were computed using the F-test and the Breusch-Pagan Lagrange Multiplier (LM) test respectively to ensure that the ordinary least squares (OLS) model is not the best model.
$$
{\rm Test \ 1:} \ H_0: {\rm OLS \ model \ is \ a \ better \ fit \ \ vs \ } H_a: {\rm Fixed-effect \ model \ is \ a \ better \ fit}
$$
$$
{\rm Test \ 2:} \ H_0: {\rm OLS \ model \ is \ a \ better \ fit \ \ vs \ } H_a: {\rm Random-effect \ model \ is \ a \ better \ fit}
$$
As both p-values were close to 0, we conclude that the OLS model is not a good fit for our data. We then moved on to check whether the fixed-effect model or the random-effect model gives us a better fit using Hausman Test with the following hypothesis:
$$
H_0: {\rm Random-effects \ model \ is \ a \ better \ fit \ \ vs \ } H_a: {\rm Fixed-effects \ model \ is \ a \ better \ fit}
$$
Here we obtained a p-value of 0.68, suggesting that $H_0$ cannot be rejected and therefore a random effects model is a better fit. However, it is important to note that setting countries as random entities would reduce the explainability of our current model.

To check for (2), we once again used the Breusch-Pagan LM test with the following hypothesis:
$$
H_0: {\rm Time-effects \ is \ significant \ \ vs \ } H_a: {\rm Time-effects \ is \ insignificant}
$$
The p-value obtained in this test is less than 0.002. Using a threshold $\alpha < 0.05$, we reject $H_0$ and conclude that our current model is a better fit.

Lastly, it is important to test for cross-sectional dependence (3) in our model as we previously assumed that entities within each country are not correlated. Once again, using the Breusch-Pagan LM test with the following hypothesis, we obtained an extremely small p-value which suggests that cross-sectional dependence exist within our model.
$$
H_0: {\rm No \ cross-sectional \ dependence \ \ vs \ } H_a: {\rm Cross-sectional \ dependence \ exist}
$$
Overall, we managed to satisfy most of the assumptions of a fixed-effect model, except for the fact that cross-sectional dependence exist within our model, which suggests that the predictors are not entirely independent, which can lead to Type-1 error inflation.


# Causal interpretation 

Since this study is observational, causality is difficult attain directly due to selection bias. Therefore, propensity score analysis was preformed using the lag of the stay-at-home variable as the treatment variable. The variables selected for the causal inference were the 14-day lagged effect of stay-at-home order, 14-day lagged effect of change in stay-at-home order, and 14-day lagged effect of change in rate of active infections as these three variables were significant in predicting change in rate of active infections. However, lag of stay at home is not a binomial variable as it has 3 levels (0,1, and 2). Therefore, it was appropriate to develop the propensity scores using a Generalized Boosted Model [12]. 


```{r, results='hide', warning=FALSE, include=FALSE}
library(twang)
mnps.AOD<-mnps(lag_stay_home ~ lag_stay_home_diff_07+lag_delta_infec_07ma, data=covid_model.d7)
```


```{r}
library(survey)
covid_model.d7$weight<- get.weights(mnps.AOD)
model_new = plm(delta_infec_07ma ~ lag_stay_home + lag_stay_home_diff_07 +lag_delta_infec_07ma, data = covid_model.d7, weights = weight,
                index = c("Country", "Date_reported"), model = "within")
nn = as.data.frame(summary(model_new)$coef)
rownames(nn) = c("Stay-home restriction at t-14 == 1",
                 "Stay-home restriction at t-14 == 2",
                 "Change in stay-home restriction at t-14 == +2",
                 "Change in stay-home restriction at t-14 == -1",
                 "Change in stay-home restriction at t-14 == +1",
                 "Change in rate of active infections at t-14")
nn
```

Once the model with the propensity score weighted results was created, it was found that no significant changes were made to the model meaning that the estimates were similar, and variables were significant in both models (shown above). This leads us to conclude that causal inference can be applied which means that when the stay-at-home mandate is at level 1 or level 2 then there is a decrease in change in rate of active infections. 

# Discussion 

## Recap of Project
The primary goals of this project is to see the effect of stay at home mandates on rate of infections for the COVID-19 virus as well as to see if there is a lag component of the mandate on rate of infections. Furthermore, the goal is to build a model to predict change in rate of active infections. The data used for analysis was downloaded from World Health Organization. Variables retained for analysis were daily frequency of new cases, new deaths, as well as cumulative cases and cumulative deaths. Data pertaining to stay-at-home mandate was downloaded from the ‘Our World in Data’ database.  The basic unit of analysis for this dataset are the daily frequency of new cases for the period October 1st to November 30th for each of these five countries (United Kingdom, Germany, Poland, Italy, France). Selection of the time-period as well as region was done to control for the effect of demographics like race and ethnicity, conflicts in implementation of mask mandates between local and national governments, effect of mutations of the COVID-19 virus as well as the effect of the COVID-19 vaccine on the frequency of daily new cases. <br>

The variables were plotted against time by country to see their distributions. The key indicator of the exploratory data analysis was the Daily Percent change in infectious population. From the result, it is seen that each of the five countries had an increasing percent change at the beginning of the time interval (October 1st), but over time the daily percent change turned negative. This distribution was the same for all five countries and possibly indicates a negative association between stay-at-home mandate and infectious population. 

## Findings of Inferential Analysis
Panel regression analysis was used to build a predictive model for the response variable change in rate of active infections by country. The variables in the model were Country (time invariant fixed factor), stay-at-home mandate. Lag variables were created based on number of days prior to or after the mandate was implemented and this lag variable was dummy coded. Multiple models were considered to find the best model for the outcome variable. The models that were considered for the analysis were the fixed effects model, an ordinary least squares model, a random effects model, and a time variant model. Multiple tests like the F-test, Hausman test, and Lagrange Multiplier Test were used to select the best predictive model for rate of active infections. The results of the analyses indicate that the fixed effects model was the best at predicting change in rate of active infections. The inferential analysis findings were that 14-day lagged effect of stay-at-home order, 14-day lagged effect of change in stay-at-home order, and 14-day lagged effect of change in rate of active infections showed statistically significant associations with change in rate of active infections. Mandate recommending not leaving the house (level 1) was negatively associated with change in rate of active infections when compared to no-restrictions (level 0) on stay-at-home mandate.  Mandate requiring not leaving the house with exceptions to daily exercise, grocery shopping and essential trips (level 2) was also negatively associated with change in rate of active infections when compared to no-restrictions (level 0) on stay-at-home mandate. There is also a positive association of a 14-day lag between change in stay-at-home order level and change in rate of active infections for all levels (no restriction; requiring not leaving the house with exceptions to daily exercise, grocery shopping and essential trips; and no restriction to recommending not leaving the house). Rate of active infections was positively associated at a prior lag of 14-days to rate of active infections at lag 0.  <br>

Based on propensity scores from causal inference, we can conclude that in the selected countries of Europe, stay-at-home policies can predict significant changes in rate of infections. <br> 

## Caveats of Current Analysis
There were a few caveats to this study. The first is that the data is only taken from October 1st to November 30th, which does not allow us to account for the effect of different seasons on rate of active infections. The second caveat is the limited generalization of result findings. To control for comorbidities the data were selected from just 5 European countries. Generalizations of the results is, therefore, limited to the selected countries.  The final caveat is that the Breusch-Pagan LM test showed that cross-sectional dependence exists within the model meaning that change in rate of active infections was not independent which could possibly lead to type 1 error inflation. <br>

## Suggestions for Research
Future Research in this area could include the selection of a more representative sample from around the world as well as running the analysis for a longer time-period to account for changing seasons. Comorbidities, age, and race have been found to be positively associated with COVID-19. Future research should include these variables and run the analysis at the individual level while controlling for the clustering effect of region.  Research can also be branched to include other Covid-19 policies such as the mask mandate and the effect of the different Covid-19 vaccines on rate of infections. 


# Acknowledgement {-}

Working codes can be found at: https://github.com/yixlu/Covid_project

# Reference {-}
[1] Fowler, J., Hill, S., Levin, R., Obradovich, N..(2020) The Effect of Stay-at-Home Orders on COVID-19 Infections on the United States. [medrxiv.org](https://www.medrxiv.org/content/10.1101/2020.04.13.20063628v2.full) <br>
[2] Torres-Reyna, O. (2010) Getting Started in Fixed/Random Effects Models using R. Lecture slides. [dss.princeton.edu](https://dss.princeton.edu/training/Panel101R.pdf) <br>
[3] Schmidheiny, K. (2020) Panel Data: Fixed and Random Effects. Lecture Notes. [schmidheiny.name](https://www.schmidheiny.name/teaching/panel2up.pdf)
<br>
[4] Thomas Hale, Sam Webster, Anna Petherick, Toby Phillips, and Beatriz Kira (2020). Oxford COVID-19 Government Response Tracker, Blavatnik School of Government. <br>
[5] Fowler, J. H., Hill, S. J., Levin, R., & Obradovich, N. (2020). The effect of stay-at-home orders on COVID-19 infections in the United States. arXiv preprint arXiv:2004.06098. <br>
[6]	European countries by Population (2021). (n.d.). Retrieved March 05, 2021, from https://www.worldometers.info/population/countries-in-europe-by-population/ <br>
[7]	CDC (11 February 2020). "Coronavirus Disease 2019 (COVID-19)". Centers for Disease Control and Prevention. Retrieved 6 December 2020. <br>
[8]	Gandhi RT, Lynch JB, Del Rio C (April 2020). "Mild or Moderate Covid-19". The New England Journal of Medicine. 383 (18): 1757–1766. doi:10.1056/NEJMcp2009249. PMID 32329974 <br>
[9]	Ritchie, R. (n.d.). Covid-19: Stay-at-home restrictions. Retrieved March 05, 2021, from https://ourworldindata.org/covid-stay-home-restrictions <br>
[10]	Ethnicity and race by countries. (n.d.). Retrieved March 05, 2021, from https://www.infoplease.com/world/social-statistics/ethnicity-and-race-countries <br>
[11]	Risk for Covid-19 infection, hospitalization, and death by race/ethnicity. (n.d.). Retrieved March 05, 2021, from https://www.cdc.gov/coronavirus/2019-ncov/covid-data/investigations-discovery/hospitalization-death-by-race-ethnicity.html <br>
[12]	Olmos, A., & Govindasamy, P. (2015). A Practical Guide for Using Propensity Score Weighting in R. Practical Assessment, Research & Evaluation, 20(13). Available online: http://pareonline.net/getvn.asp?v=20&n=13 <br>



# Session info {-}

<span style='color:blue'>
Report information of your `R` session for reproducibility. 
</span> 


```{r}
sessionInfo()
```
